---
title: "Superfast PRS"
author: "Julian Echave"
date: "5/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r}
# Load libraries and functions
library(here)
library(bio3d)
library(tidyverse)
library(cowplot)
library(penm)
library(jefuns)
library(Matrix)

library(patchwork)

library(microbenchmark) # benchmarking times
library(profvis) # profiling
```

```{r}
# Set parameters
pdb_id = "2acy_A"
# set wt
pdb <- read.pdb(here("data-raw/2acy_A.pdb"))
wt <- set_enm(pdb,  "ca",  "anm",  10.5,  F)
mut <- get_mutant_site(wt, 80, 1)
```


## temp tests
```{r}
nmut <- 100
mut_dl_sigma <- 0.3
mut_sd_min <- 1
tic()
prs <- prs_dr2ij_smrs(wt, nmut, "lfenm", mut_dl_sigma, mut_sd_min, 1024)
toc()
tic()
dms <- dmsmat <- dmsmat_dms.new(wt, nmut, mut_dl_sigma, mut_sd_min, 1024)
toc()


```









## compare values






## compare cpu times

mrs_all: simulation-based prs in which first a set of mutants is generated, then each wt vs mut pair comparison is performed before averaging over mutants.

smrs_all:  instead of looping over mutations at a given site, smrs_all calculates an fmat with nmut columns for each site. Thus f(i,j, m) is the force at i due to mutation m at a site j.

amrs: analytic averaging over mutations, thus avods generating individual mutants.



```{r}
microbenchmark(mrs_all(wt, 10, "lfenm", 0.3, 1, 1024),
               smrs_all(wt, 10, "lfenm", 0.3, 1, 1024),
               amrs_all(wt, 0.3, 1),
               times = 1)
```


```{r}
mutants <- generate_mutants(wt, 20, "lfenm", 0.3, 1, 1024)
perturbations <- generate_perturbations(wt, 20, 0.3, 1, 1024)
microbenchmark(calculate_dr2ij_mrs(mutants),
               calculate_dr2ij_smrs(wt, perturbations$fmat),
               calculate_dr2ij_amrs(wt, 0.3, 1),
               times = 1)
```
```{r}

microbenchmark(calculate_df2ij_mrs(mutants),
               calculate_df2ij_smrs(wt, perturbations$fmat),
               calculate_df2ij_amrs(wt, 0.3, 1),
               times = 1)
```
```{r}
microbenchmark(calculate_de2ij_mrs(mutants),
               calculate_de2ij_smrs(wt, perturbations$fmat),
               calculate_de2ij_amrs(wt, 0.3, 1),
               times = 1)
```

```{r}
microbenchmark(calculate_dvsij_mrs(mutants),
               calculate_dvsij_smrs(wt, perturbations$dlmat),
               calculate_dvsij_amrs(wt, 0.3, 1),
               times = 1)
```

## Comparison of results

mrs_all and smrs_all give identical results which differ from amrs because of convergence (prs_amrs is the asymptotic "true" average).



```{r}
dat.sim <- mrs_all(wt, 20, "lfenm", 0.3, 2, 1024)  %>%
  pivot_longer(cols = df2ij:de2ij,
               names_to = "response_type",
               values_to = "value")

dat.new <- smrs_all(wt, 20, "lfenm", 0.3, 2, 1024) %>%
  pivot_longer(cols = df2ij:de2ij,
               names_to = "response_type",
               values_to = "value")

dat_amrs <- prs_amrs(wt, 0.3, 2) %>%
  select(i, j, df2ij, dr2ij, de2ij) %>%
  pivot_longer(cols = df2ij:de2ij,
               names_to = "response_type",
               values_to = "value")

dat <- inner_join(dat.new, dat_amrs, by = c("i", "j", "response_type"), suffix = c(".new", "_amrs"))

p1 <- dat %>%
  ggplot(aes(value_amrs, value.new)) +
  geom_point() +
  geom_abline() +
  facet_wrap(~response_type, ncol = 3, scales = "free") +
  ggtitle("new vs fast")

dat <- inner_join(dat.new, dat.sim, by = c("i", "j", "response_type"), suffix = c(".new", ".sim"))

p2 <- dat %>%
  ggplot(aes(value.sim, value.new)) +
  geom_point() +
  geom_abline() +
  facet_wrap(~response_type, ncol = 3, scales = "free") +
  ggtitle("new vs sim")


plot_grid(p1, p2, ncol = 1)



```
