---
title: "Perturbation Response Scanning using penm"
author: "Julian Echave"
date: "5/12/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r}
# Load libraries and functions

library(tidyverse)
library(ggridges)
library(cowplot)
library(bio3d)
library(here)
library(penm)
library(jefuns)

# source functions to use in this script
source("lfenm_plot.R")
```

```{r}
# Set parameters
pdb_id = "2acy_A"
```







# Structural response theory

## Structural response, definitions

$f_{ij} = k_{ij} \delta l_{ij}$ is the force in the direction of contact $i-j$.

Definitions:

$$ de = K^{1/2}dr = C^{-1/2}df \\ dr = C df $$

## Structural response, site and mode respresentations
Therefore, in normal-mode representation: 

$$de_{n} = \sigma_n df_{n} \\ dr_n = \sigma_n^2 df_n$$.

By analogy, in site representation, check if it's approximately so that 

$$de^2_{i} = \sigma_i^2 df^2_{i} \\ dr^2_i = \sigma_i^4 df^2_i$$.

## Continuation

Remember that, on average: 

$$df^2_n \propto \frac{1}{\sigma_n^2}$$ 

and probably 

$$df^2_i \propto 1 / \sigma_i^2$$




# Perturbation Response Scanning

```{r}
# Prepare site-site response data

dfij <- here("data-raw", paste0(pdb_id, "_dfij.csv")) %>% 
  read_csv()

### Make long-version of dfij


dfij_long <- dfij  %>% 
  pivot_longer(cols = c(dr2ij, de2ij, df2ij),
               names_to = "response",
               values_to = "value") %>% 
  mutate(response = factor(response, levels = c("df2ij", "de2ij", "dr2ij"))) 

dfij_long

```


## Site-site average response matrices


```{r}
dfij_long %>%
  group_by(i, j, response) %>%
  summarise(value = mean(value)) %>% 
  ungroup() %>% 
  # filter(abs(j - i) >= 1) %>% 
  group_by(response) %>% 
  mutate(value = mMnorm(value)) %>% 
  ggplot(aes(j, i, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_cowplot() +
  theme(legend.position = "none") +
  coord_fixed() +
  facet_grid(. ~ response) +
  NULL 
```

## Response matrices are not symmetric


```{r}
plot_mji_vs_mij <- function(m, title = "") {
  tm = t(m)
  df <- matrix_to_tibble(m)
  dft <- matrix_to_tibble(tm)
  df$mji = dft$mij
  
  df %>% 
    ggplot(aes(mij, mji)) +
    geom_point(size = .3, alpha = .3) +
    geom_smooth(method = "lm") +
    geom_abline() +
    ggtitle(title) +
    coord_fixed() +
    theme_cowplot()
}


#dr2ij asymmetry
df <- dfij %>%
  group_by(i, j) %>%
  summarise(dr2ij = mean(dr2ij),
            de2ij = mean(de2ij),
            df2ij = mean(df2ij)) %>% 
  arrange(j, i)

nsites <- get_nsites(wt)

p1 <- plot_mji_vs_mij(offdiag(matrix(df$df2ij, nsites, nsites)), "df2ij") 
p2 <- plot_mji_vs_mij(offdiag(matrix(df$de2ij, nsites, nsites)), "de2ij") 
p3 <- plot_mji_vs_mij(offdiag(matrix(df$dr2ij, nsites, nsites)), "dr2ij") 

plot_grid(p1, p2, p3, ncol = 3)
```


## Response over a single scan is similar to average over scans

```{r}
dfij_long %>%
  filter(mutation %in% c(1, 2)) %>% 
  group_by(mutation, response) %>% 
  mutate(value = mMnorm(value)) %>% 
  ggplot(aes(j, i, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_cowplot() +
  theme(legend.position = "none") +
  coord_fixed() +
  facet_grid(mutation ~ response) +
  NULL 
```

## Response over single scan similar to average, cont.
```{r}
dfij_long %>% 
  filter(mutation %in% c(1, 2)) %>% 
  pivot_wider(names_from = mutation,
              names_prefix = "mutation_",
              values_from = value)  %>% 
  ggplot(aes(mutation_1, mutation_2)) +
  geom_point() +
  geom_smooth() +
  geom_abline() +
  theme_cowplot() +
  facet_wrap(~ response, ncol = 3, scales = "free")
```


## sd(response) is proportional to mean(response)

```{r}
dfij_long %>% 
  group_by(i, j, response) %>% 
  mutate(value_mean = mean(value),
         value_sd = sd(value)) %>% 
  ggplot(aes(value_mean, value_sd)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_cowplot() +
  facet_wrap(~ response, scales = "free")
```


# Average response of site i


## Average response vs. site

```{r}
dfij_long %>% 
  group_by(i, msfi, response) %>% 
  summarise(value = mean(value)) %>% 
  ggplot(aes(i, value)) +
  geom_line() +
  theme_cowplot() +
  facet_grid(response ~ ., scales = "free_y") +
  ggtitle("response_of(i) vs. site")
```

## Average response vs. msf

```{r}
dfij_long %>% 
  group_by(i, msfi, response) %>% 
  summarise(value = mean(value)) %>% 
  ggplot(aes(msfi, value)) +
  geom_point() +
  geom_smooth() +
  theme_cowplot() +
  facet_grid(response ~ ., scales = "free_y") +
  ggtitle("response_of(i) vs. msf(i)")
```

## Average response vs. local packing density

```{r}
dfij_long %>% 
  group_by(i, mlmsi, response) %>% 
  summarise(value = mean(value)) %>% 
  ggplot(aes(mlmsi, value)) +
  geom_point() +
  geom_smooth() +
  theme_cowplot() +
  facet_grid(response ~ ., scales = "free_y") +
  ggtitle("response_of(i) vs. cn(i)")
```


# Influence of site j

## Average response vs. mutated site j

```{r}
dfij_long %>% 
  group_by(j, msfj, response) %>% 
  summarise(value = mean(value)) %>% 
  ggplot(aes(j, value)) +
  geom_line() +
  theme_cowplot() +
  facet_grid(response ~ ., scales = "free_y") +
  ggtitle("response_to(j) vs. j")
```


## Average response vs. msf of mutated site j

```{r}
dfij_long %>% 
  group_by(j, msfj, response) %>% 
  summarise(value = mean(value)) %>% 
  ggplot(aes(msfj, value)) +
  geom_point() +
  geom_smooth() +
  theme_cowplot() +
  facet_grid(response ~ ., scales = "free_y") +
  ggtitle("response_of(j) vs. msf(j)")
```

## Average response vs. local packing density of mutated site j

```{r}
dfij_long %>% 
  group_by(j, mlmsj, response) %>% 
  summarise(value = mean(value)) %>% 
  ggplot(aes(mlmsj, value)) +
  geom_point() +
  geom_smooth() +
  theme_cowplot() +
  facet_grid(response ~ ., scales = "free_y") +
  ggtitle("response_of(j) vs. mlms(j)")
```


# Structural response along normal modes


```{r}
# Prepare data

dfnj <- here("data-raw", paste0(pdb_id, "_dfnj.csv")) %>% 
  read_csv()
  
dfnj_long  <- dfnj %>% 
  pivot_longer(cols = c(dr2nj, de2nj, df2nj),
               names_to = "response",
               values_to = "value") %>% 
  mutate(response = factor(response, levels = c("df2nj", "de2nj", "dr2nj"))) 

dfnj_long
```

```{r}
# prepare data
df_prs_mode <- dfnj %>%
  select(j, mutation, mode, dr2nj) %>%
  rename(dr2 = dr2nj) 
```


## Structrural response along normal modes

```{r}
pnj = df_prs_mode %>%
  group_by(mode, j) %>%
  summarise(dr2_mean = mean(dr2),
            dr2_sd = sd(dr2)) %>%
  ggplot(aes(mode, j, fill = sqrt(dr2_mean))) +
  geom_tile() +
  scale_color_viridis_c() +
  theme_cowplot()

pn <- df_prs_mode %>%
  group_by(mode) %>%
  summarise(dr2_mean = mean(dr2),
            dr2_sd = sd(dr2)) %>%
  ggplot(aes(mode, dr2_mean)) +
  geom_line() +
  theme_cowplot()

pj <- df_prs_mode %>%
  group_by(j) %>%
  summarise(dr2_mean = mean(dr2),
            dr2_sd = sd(dr2)) %>%
  ggplot(aes(j, dr2_mean)) +
  geom_line() +
  theme_cowplot()

plot_grid(pnj, plot_grid(pn, pj), ncol = 1)
```

## Single scan similar to average over scans

```{r message = F}
p1 <- df_prs_mode %>% 
  filter(mutation != 0) %>% 
  group_by(mode, mutation) %>% 
  summarise(dr2n = sum(dr2))  %>% 
  mutate(mutation = factor(mutation)) %>% 
  ggplot(aes(mode, dr2n, color = mutation)) +
  geom_line() +
  theme_cowplot() +
  theme(legend.position = c(0.1, 0.9)) +
  scale_color_viridis_d()

p2 <- df_prs_mode %>% 
  filter(mutation != 0) %>% 
  group_by(mode, mutation) %>% 
  summarise(dr2n = sum(dr2)) %>% 
  ungroup() %>% 
  group_by(mode) %>% 
  mutate(dr2n_mean = mean(dr2n)) %>% 
  mutate(mutation = factor(mutation)) %>% 
  ggplot(aes(dr2n_mean, dr2n, color = mutation)) +
  geom_point() +
  geom_smooth() +
  scale_color_viridis_d() +
  theme_cowplot() +
  theme(legend.position = c(0.1, 0.9)) +
  NULL


plot_grid(p1, p2)

```

