---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and knit. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# penm

<!-- badges: start -->
[![R-CMD-check](https://github.com/jechave/penm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jechave/penm/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`penm` builds Elastic Network Models (ENMs) of proteins, perturbs them, and
measures how far the perturbed protein has moved from the original in energy,
structure and motion.

An ENM represents a protein as beads joined by springs: one node per residue,
a spring between every pair closer than a cutoff. Diagonalising the resulting
network gives the normal modes — the protein's soft collective motions. `penm`
then *perturbs* that network, by default by mutating a site, and gives you a
menu of measures of the difference between the original and the perturbed
protein, resolved either by site or by normal mode.

The name means "perturb ENM": mutation is the perturbation implemented today,
and the design allows for others.

## Installation

```r
# install.packages("remotes")
remotes::install_github("jechave/penm")
```

## Usage

Everything starts from `set_enm()`, which turns a `bio3d` pdb object into a
`prot`. The package bundles one, `pdb_2acy_A`, so this runs as-is; for your own
structure use `bio3d::read.pdb("your.pdb")` (and `bio3d::trim.pdb()` to pick a
chain).

```{r example}
library(penm)

wt <- set_enm(pdb_2acy_A, node = "ca", model = "ming_wall",
              d_max = 10.5, frustrated = FALSE)

get_nsites(wt)   # number of nodes
get_nmodes(wt)   # 3 * nsites - 6
```

Mutate a site to get a second `prot`:

```{r mutate}
mut <- get_mutant_site(wt, site_mut = 11, mutation = 1, ensemble = 7)
```

Then measure the difference. The `delta_*` families each take the pair, and
report either a scalar, a profile over sites, or a profile over modes:

```{r measure}
ddg_dv(wt, mut)                            # scalar: minimum-energy difference

dr2i <- delta_structure_dr2i(wt, mut)      # one value per site
which.max(dr2i)                            # the site that moved most
```

`wt` and `mut` are *roles in a comparison*, not two kinds of object — both are
`prot` objects, and either may play either part. That is what makes an
evolutionary trajectory a loop: feed each mutant back as the next generation's
wild type, and strain accumulates.

```{r trajectory}
p <- wt
for (gen in 1:3) {
  p <- get_mutant_site(p, site_mut = 10 * gen, mutation = 1, ensemble = 7)
  cat("generation", gen, "  minimum energy:", enm_v_min(p), "\n")
}
```

## Learn more

- `vignette("penm")` — the full build → mutate → measure walkthrough.
- `?penm` — the function map.
- `?penm_ensemble` — what `ensemble` means, and when it should change.

## References

The model and its derivation:

- Echave J (2008). Evolutionary divergence of protein structure: the linearly
  forced elastic network model. *Chemical Physics Letters* **457**(4–6),
  413–416. <doi:10.1016/j.cplett.2008.04.042>
- Echave J, Fernández FM (2010). A perturbative view of protein structural
  variation. *Proteins* **78**(1), 173–180. <doi:10.1002/prot.22553>

Run `citation("penm")` for the same in BibTeX.
