---
title: "Effect of mutating the ENM Kmat on motions"
author: "Julian Echave"
date: "4/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load libraries and functions

```{r}
# load libraries
library(tidyverse)
library(ggridges)
library(cowplot)
library(bio3d)
library(penm)
library(jefuns)

# source functions to use in this script
source("mutant_scan_functions.R")
source("lfenm_plot.R")

# set paths and file names

pdb_path <- "data" # pdb files repaired using FoldX (by Elisha)

```



# Set up parameters

```{r}
# global parameters
beta <- beta_boltzmann()
# other parameters

param <- lst(
  enm = lst(model = "ming_wall",
            d_max = 10.5,
            frustrated = NA),
  mut = lst(model = "lfenm",
            dl_sigma = 0.3,
            mut_sd_min = 1,
            update_enm = NA),
  prs = lst(nmut_per_site = 1),
)

str(param)

```


# Read and preform ENM NMA of wild-type protein

## Check pdb file and read wt coordinates

```{r}
# set pdb data
pdb <- "2ACY"
chain <- "A"
pdb_prefix <- "RepairPDB_"

# perform various checks of file existence and consistence
check <- pdb_file_check(pdb, chain, pdb_path, pdb_prefix)

# set up wild-type
wt <- read_pdb_sc(pdb, chain, pdb_path, pdb_prefix)

str(wt)

```


## Initialize wild-type

Check that it works when pdb_active_site is defined
```{r}
dummy_active_site = wt$pdb_site[5:7]
wt <- init_prot(prot = wt,
                pdb_site_active = dummy_active_site,
                ideal = wt, # look what is this ideal for (frustrated?)
                model = param$enm$model,
                d_max = param$enm$d_max,
                frustrated = T)

str(wt)
```


## ENM mutations chnage entropic contributions

Check that entropic tontributions to energy change when update_enm = TRUE

```{r}
df <- tibble(prot = c("wt", "mutF", "mutT"),
             site = c(3, 3, 3),
             wt = lst(wt),
             mutation = c(0, 1, 1),
             update_enm = c(F, F, T),
             frustrated = c(F, F, F)) %>%
  mutate(mut_energy = pmap(list(wt, site, mutation, update_enm, frustrated), get_mut_energy)) %>%
  unnest(mut_energy) %>%
  arrange(mutation, frustrated, update_enm) %>%
  select(-prot, -wt, -mutation)

df

```

## Check effect of adding frustrations

Adding frustrations affects entropic terms, thus only when updating ENM

```{r}
# with and without frustration

df <- tibble(prot = c("wt", "mutff", "mutft", "muttf", "muttt"),
             site = rep(80, 5),
             wt = lst(wt),
             mutation = c(0, 1, 1, 1, 1),
             update_enm = c(F, F, F, T, T),
             frustrated = c(F, F, T, F, T)) %>%
  mutate(mut_energy = pmap(list(wt, site, mutation, update_enm, frustrated), get_mut_energy)) %>%
  unnest(mut_energy) %>%
  arrange(mutation, update_enm, frustrated) %>%
  select(-prot, -wt, -mutation)

df

```

## Plot change in msf profile

```{r}

# get mutant
mutated_site <- 80
update_enm <- T
frustrated <- F
mut <- get_mut(wt, mutated_site, mutation = 1, dl_sigma = 1, update_enm, frustrated)

# set msf data.frame for plotting
site <- get_site(wt)
wt_msf <- get_msf_site(wt)
mut_msf <- get_msf_site(mut)
df <- tibble(site, wt_msf, mut_msf)

p1  = df %>%
  pivot_longer(cols = c("wt_msf", "mut_msf"),
               names_to = "protein",
               values_to = "msf") %>%
  ggplot(aes(site, msf, color = protein)) +
  geom_line() +
  scale_color_viridis_d() +
  theme_cowplot() +
  theme(axis.text.y = element_blank()) +
  theme(legend.position = "top") +
  NULL +
  ggtitle(paste0("mutated_site:", mutated_site, ", update_enm = ", update_enm,
                 ", frustrated = ", frustrated))

p2 <- df %>%
  ggplot(aes(site, mut_msf - wt_msf)) +
  geom_line() +
  theme_cowplot()  +
  geom_hline(yintercept = 0) +
  theme(axis.text.y = element_blank()) +
  NULL

plot_grid(p1, p2, ncol = 1)
```





## Change of msf along normal modes

```{r}

# get mutant
mutated_site <- 80
update_enm <- T
frustrated <- F
mut <- get_mut(wt, mutated_site, mutation = 1, dl_sigma = 1, update_enm, frustrated)

# set msf data.frame for plotting
mode <- get_mode(wt)
wt_msf <- 1 / get_evalue(wt)
mut_msf <- 1 / get_evalue(mut)
df <- tibble(mode, wt_msf, mut_msf)

p1  = df %>%
  pivot_longer(cols = c("wt_msf", "mut_msf"),
               names_to = "protein",
               values_to = "msf") %>%
  ggplot(aes(mode, msf, color = protein)) +
  geom_line() +
  scale_color_viridis_d() +
  theme_cowplot() +
  theme(axis.text.y = element_blank()) +
  theme(legend.position = "top") +
  NULL +
  ggtitle(paste0("mutated_site:", mutated_site, ", update_enm = ", update_enm,
                 ", frustrated = ", frustrated))

p2 <- df %>%
  ggplot(aes(mode, mut_msf - wt_msf)) +
  geom_line() +
  theme_cowplot()  +
  geom_hline(yintercept = 0) +
  theme(axis.text.y = element_blank()) +
  NULL

plot_grid(p1, p2, ncol = 1)
```


```{r}
omn <- t(wt$enm$umat) %*% mut$enm$umat

nmodes <- nrow(omn)

matrix_to_tibble <- function(mat) {
  M <- nrow(mat)
  N <- ncol(mat)
  i <- rep(seq(N), M)
  j <- rep(seq(M), each = N)
  df <- tibble(i, j, mat_ij = as.vector(mat))
  df
}

plot_matrix <- function(mat) {
  df <- matrix_to_tibble(mat)
  plot <-  df %>%
    ggplot(aes(i, j, fill = mat_ij)) +
    geom_tile() +
    scale_fill_viridis_c() +
    theme_cowplot() +
    NULL +
    # theme(legend.position = "none") +
    coord_fixed() +
    NULL
  plot
}

plot_matrix(omn) +
  xlab("n_wt") +
  ylab("n_mut")

plot(diag(omn^2))
```
