---
title: "Getting started with penm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with penm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r setup}
library(penm)
```

## The model

An Elastic Network Model (ENM) represents a protein as nodes joined by springs:
one node per residue, and a spring between every pair of nodes closer than a
cutoff $d_{max}$ in the native structure. For a conformation $\mathbf{r}$, the
potential energy is

$$V(\mathbf{r}) = \sum_{ij} \tfrac{1}{2} k_{ij}\,(r_{ij} - l_{ij})^2 ,$$

with $r_{ij}$ the distance between nodes $i$ and $j$ in $\mathbf{r}$, $l_{ij}$
the spring's rest length, and $k_{ij}$ its force constant.

Expanding $V$ to second order about the equilibrium structure gives the
stiffness matrix $\mathbf{K}$, the Hessian of $V$. Its eigenvectors are the
**normal modes** and its eigenvalues $\lambda_n$ their force constants: mode $n$
has mean-square amplitude $\propto 1/\lambda_n$, so the smallest eigenvalues
carry the largest fluctuations. Six eigenvalues vanish, corresponding to the
rigid-body translations and rotations, leaving $3N - 6$ internal modes.

Given such a network, `penm` — "perturbed ENM" — builds the network of a mutant.
Mutations follow the linearly forced ENM (LFENM) of Echave (2008) and Echave &
Fernández (2010): mutating a site shifts the rest lengths $l_{ij}$ of the springs
connected to it, and the structure relaxes to a new equilibrium.

The mutant is a protein like any other in this package. Its structure, energy and
modes are computed by the same functions as the wild type's, the two can be
compared, and it can be mutated in turn — which is what the last two sections
build on.

## Building a network

`set_enm()` builds the network from a `pdb` object, as returned by
`bio3d::read.pdb()`, and returns a `prot`. Every other function takes that
`prot`.

The examples below use `pdb_2acy_A`, a copy of acylphosphatase (98 residues,
chain A) bundled with the package.

```{r build}
wt <- set_enm(pdb_2acy_A, node = "ca", model = "ming_wall",
              d_max = 10.5)

get_nsites(wt)
get_nmodes(wt)
```

### Using your own structure

Any structure `bio3d` can read will do. Select a single chain unless you intend
a multi-chain network, since chains that make no contacts give disconnected
components:

```r
library(bio3d)

pdb <- read.pdb("path/to/your/structure.pdb")   # or read.pdb("2acy") to fetch
pdb_A <- trim.pdb(pdb, chain = "A")

wt <- set_enm(pdb_A, node = "ca", model = "ming_wall",
              d_max = 10.5)
```

All four arguments are required, so that the network is recorded in the call:

- **`node`** — which atom represents a residue: `"ca"` (alpha carbons), `"sc"`
  (side-chain centroids) or `"cb"` (beta carbons).
- **`model`** — the spring-constant function. `"anm"` gives every contact the
  same constant; `"ming_wall"` (used here), `"hnm"`, `"hnm0"`, `"pfanm"` and
  `"reach"` make $k_{ij}$ depend on distance or on separation along the sequence.
- **`d_max`** — the contact cutoff in ångström. For Cα networks, 10–12 Å is the
  usual range. Shorter cutoffs give sparser, softer networks.

These choices define the network, so hold them fixed across any comparison.

## What a `prot` holds

Accessors read the network: `get_xyz()` the coordinates, `get_kmat()` the
stiffness matrix, `get_evalue()` and `get_umat()` the modes, `get_msf_site()`
the fluctuation of each site. `?get_prot_property` lists them all; use them
rather than the object's internals, which are free to change.

Two views of the same network. The eigenvalue spectrum, spanning nearly three
orders of magnitude from the softest mode to the stiffest:

```{r spectrum, eval = has_ggplot, fig.width = 7, fig.height = 3}
library(ggplot2)

ggplot(tibble::tibble(mode = get_mode(wt), evalue = get_evalue(wt)),
       aes(mode, evalue)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "mode", y = "eigenvalue") +
  theme_minimal()
```

And the fluctuations those modes imply, site by site — the exposed loops move,
the buried core does not:

```{r msf, eval = has_ggplot, fig.width = 7, fig.height = 3}
ggplot(tibble::tibble(site = get_site(wt), msf = get_msf_site(wt)),
       aes(site, msf)) +
  geom_line() +
  labs(x = "site", y = "mean-square fluctuation") +
  theme_minimal()
```

A network from `set_enm()` sits at its own minimum: every spring is at its rest
length, so there is no strain.

```{r unstrained}
enm_v_min(wt)
```

## Mutating

`get_mutant_site()` perturbs the springs of one site and relaxes the structure,
returning a network of the same kind as its input:

```{r mutate}
mut <- get_mutant_site(wt, site_mut = 11, mutation = 1, ensemble = 1)

c(wt = get_nsites(wt),  mut = get_nsites(mut))
c(wt = enm_v_min(wt),   mut = enm_v_min(mut))
```

Every accessor and profile above applies to `mut` as it does to `wt`. Only the
values differ — here the mutant carries strain the wild type does not.

`site_mut` is a sequential index from 1 to `get_nsites()`, not a PDB residue
number. The two differ whenever the file's numbering has an offset or gaps, so
convert:

```{r site-lookup}
which(get_pdb_site(wt) == 11)
```

The model has no amino acids, so a mutation is not a substitution of one residue
type for another: it is a random shift of the rest lengths $l_{ij}$ at the
mutated site, drawn with standard deviation `mut_dl_sigma` (default 0.3 Å).
`mut_sd_min` (default 2) excludes springs between residues close in sequence.
Shifting the rest lengths is what leaves the relaxed structure strained.

### Identifying a mutant

Three integers name a mutant. `site_mut` is the site. `mutation` indexes
independent draws at that site — `1`, `2`, `3` are three different mutations of
the same residue, and `0` returns the wild type unchanged. `ensemble` indexes
realizations of the whole mutational process.

The triple is hashed to seed the draw, so the same triple always yields the same
mutant: these calls are reproducible, not random from call to call.

`mutation` and `ensemble` differ in scope. Vary `mutation` for replicates within
one study; change `ensemble` to repeat an entire study — every site, every
replicate — as an independent realization. Hold `ensemble` fixed across a scan or
a trajectory. `?penm_ensemble` is the full account.

Replicates at one site show how wide the distribution of mutational effects is:

```{r replicates}
ddg <- numeric(5)
for (m in 1:5) {
  mut_m <- get_mutant_site(wt, site_mut = 11, mutation = m, ensemble = 1)
  ddg[m] <- ddg_dv(wt, mut_m)
}
ddg
```

## Measuring mutational changes

Three families measure wild-type-vs-mutant differences, all taking the pair.
Within the structure and motion families, the suffix `i` gives a profile over
sites and `n` a profile over modes.

**Energy.** Scalars. `ddg_dv` is the change in minimum energy,
$V_{min}(\text{mut}) - V_{min}(\text{wt})$; `ddg_tds` the entropic term. `penm`
reports the components separately rather than a single free energy.

```{r energy}
ddg_dv(wt, mut)
```

Since the wild type here is unstrained, this equals the mutant's strain energy.

**Structure.** Squared displacements, in Å², resolved over sites or over modes.
The two are the same deformation in different bases, and sum to the same total:

```{r structure}
dr2i <- delta_structure_dr2i(wt, mut)   # per site
dr2n <- delta_structure_dr2n(wt, mut)   # per mode

all.equal(sum(dr2i), sum(dr2n))
```

By site, the deformation is sharply localised: the mutated site moves most, and
the response falls by three orders of magnitude across the protein.

```{r dr2i-plot, eval = has_ggplot, fig.width = 7, fig.height = 3}
ggplot(tibble::tibble(site = get_site(wt), dr2 = dr2i), aes(site, dr2)) +
  geom_line() +
  geom_vline(xintercept = 11, linetype = "dashed", colour = "grey50") +
  scale_y_log10() +
  labs(x = "site", y = expression(dr^2),
       subtitle = "dashed line: the mutated site") +
  theme_minimal()
```

By mode, it is concentrated at the soft end of the spectrum. Mutations push the
structure along the motions it was already most willing to make:

```{r dr2n-plot, eval = has_ggplot, fig.width = 7, fig.height = 3}
ggplot(tibble::tibble(mode = get_mode(wt), dr2 = dr2n), aes(mode, dr2)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "mode (softest first)", y = expression(dr^2)) +
  theme_minimal()
```

```{r soft-modes}
sum(sort(dr2n, decreasing = TRUE)[1:10]) / sum(dr2n)   # share taken by 10 modes
cor(log(dr2n), log(get_evalue(wt)))
```

The two views answer different questions — where the protein moved, and which of
its motions were excited.

**Motion.** Changes in the fluctuations themselves — `delta_motion_dmsfi` over
sites, `delta_motion_dmsfn` over modes:

```{r motion}
sum(delta_motion_dmsfi(wt, mut))
```

The zero is a consequence of the perturbation model, explained next.

Each family has further members, weighting the same comparisons differently; see
`?delta_structure_by_site`, `?delta_motion_by_site` and their by-mode
counterparts.

## What the perturbation touches

Under the LFENM — `get_mutant_site()`'s default `mut_model = "lfenm"` — the
perturbation is linear: it shifts rest lengths and moves the equilibrium
coordinates, but leaves $\mathbf{K}$ unchanged. The mutant therefore has the same
normal modes and the same eigenvalues as the wild type:

```{r invariant}
all.equal(get_evalue(wt), get_evalue(mut))
all.equal(as.matrix(get_kmat(wt)), as.matrix(get_kmat(mut)))
```

Any quantity derived from the spectrum alone is consequently identical for the
two, and its difference vanishes:

```{r entropic}
ddg_tds(wt, mut)
```

This is why `delta_motion_dmsfi` returned zero above. Under this model the
informative measures are those built on coordinates.

## Influential and sensitive sites

`delta_structure_dr2i(wt, mut)` gives the response *at every site* to mutating
one site. Collecting those profiles over every mutated site in turn gives a
site-by-site matrix, whose two margins answer different questions: how much
mutating a site perturbs the rest of the protein (its **influence**), and how
much a site is perturbed by mutations elsewhere (its **sensitivity**).

A single mutant is one random draw, and draws at the same site scatter widely,
as the replicates above showed. Average over several mutants per site, or the
profiles are mostly noise:

```{r profiles}
sites <- get_site(wt)
n_mut <- 10
dr2 <- matrix(0, length(sites), length(sites))   # [responding site, mutated site]

for (s in sites) {
  acc <- numeric(length(sites))
  for (m in 1:n_mut) {
    mut_s <- get_mutant_site(wt, site_mut = s, mutation = m, ensemble = 1)
    acc <- acc + delta_structure_dr2i(wt, mut_s)
  }
  dr2[, s] <- acc / n_mut
}

influence   <- colSums(dr2)
sensitivity <- rowMeans(dr2)
```

```{r profiles-plot, eval = has_ggplot, fig.width = 7, fig.height = 3.5}
profiles <- tibble::tibble(
  site    = rep(sites, 2),
  value   = c(influence, sensitivity),
  profile = rep(c("influence", "sensitivity"), each = length(sites))
)

ggplot(profiles, aes(site, value)) +
  geom_line() +
  facet_wrap(~ profile, ncol = 1, scales = "free_y") +
  scale_y_log10() +
  labs(x = "site", y = NULL) +
  theme_minimal()
```

The two behave quite differently. Sensitivity falls off sharply with burial,
while influence is only weakly related to it:

```{r profiles-cn}
cn <- get_cn(wt)

cor(log(sensitivity), cn)
cor(log(influence), cn)
```

Sensitivity largely reflects how much a site moves to begin with, which the
unperturbed network already predicts through its mean-square fluctuation:

```{r sensitivity-msf}
cor(log(sensitivity), log(get_msf_site(wt)))
```

`get_cn()` and `get_msf_site()` are two of several single-structure profiles —
see also `get_wcn()` and `get_mlms()`.

Both profiles are comparative. Responses scale as $dr^2 \propto \sigma^2$ with
`mut_dl_sigma`, and with the spring constants, so their absolute size is
arbitrary; only the ranking of sites carries meaning.

## Evolutionary trajectories

Mutating the mutant, and so on, traces an evolutionary trajectory: a random site
each generation, the result kept as the starting point of the next. Measuring
each generation against the *original* structure tracks how far the lineage has
drifted:

```{r trajectory}
set.seed(1)
n_gen <- 30

p <- wt
traj <- tibble::tibble(gen = 1:n_gen, v_min = NA_real_, dr2 = NA_real_)

for (g in 1:n_gen) {
  site_g <- sample(get_site(wt), 1)
  p <- get_mutant_site(p, site_mut = site_g, mutation = g, ensemble = 1)

  traj$v_min[g] <- enm_v_min(p)                      # strain carried by this generation
  traj$dr2[g]   <- sum(delta_structure_dr2i(wt, p))  # divergence from the original
}

head(traj, 3)
```

```{r trajectory-plot, eval = has_ggplot, fig.width = 7, fig.height = 3.5}
ggplot(tidyr::pivot_longer(traj, c(v_min, dr2)),
       aes(gen, value)) +
  geom_line() +
  facet_wrap(~ name, ncol = 1, scales = "free_y") +
  labs(x = "generation", y = NULL) +
  theme_minimal()
```

Both grow, for different reasons. Strain accumulates because each generation
inherits the previous one's perturbed rest lengths and adds to them. Structural
divergence grows because the mutations do not undo one another — the lineage
random-walks away from its starting structure, so the drift is not monotonic
generation to generation.

Only the starting network is unstrained. Every later generation acts as the wild
type of the next comparison while itself carrying strain.

Keep `ensemble` fixed along a trajectory so the generations belong to one
realization.

## Where to go next

- `?penm` — the map of what the package offers.
- `?penm_ensemble` — the `(ensemble, site_mut, mutation)` key in depth.
- `?get_prot_property` — every accessor.
- The `delta_*` help pages — the full menu of measures.

## References

Echave J (2008). Evolutionary divergence of protein structure: the linearly
forced elastic network model. *Chemical Physics Letters* **457**(4–6), 413–416.
doi:10.1016/j.cplett.2008.04.042

Echave J, Fernández FM (2010). A perturbative view of protein structural
variation. *Proteins* **78**(1), 173–180. doi:10.1002/prot.22553
