---
title: "Perturbation Response Scanning using penm"
author: "Julian Echave"
date: "5/12/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r}
# Load libraries and functions

library(tidyverse)
library(ggridges)
library(cowplot)
library(bio3d)
library(here)
library(penm)
library(jefuns)

# source functions to use in this script
source("lfenm_plot.R")
```

```{r}
# Set parameters
pdb_id = "2acy_A"
```


# Site-site response


```{r}
# Prepare site-site response data

dfij <- here("data-raw", paste0(pdb_id, "_dfij.csv")) %>%
  read_csv()

### Make long-version of dfij


dfij_long <- dfij  %>%
  pivot_longer(cols = c(dr2ij, de2ij, df2ij),
               names_to = "response",
               values_to = "value") %>%
  mutate(response = factor(response, levels = c("df2ij", "de2ij", "dr2ij")))

dfij_long

```





```{r results = FALSE}
# Set up wt

pdb_file <- here("data-raw/2acy_A.pdb")
pdb <- read.pdb(pdb_file)

wt <- set_enm(pdb, node = "sc", model = "ming_wall", d_max = 10.5, frustrated = FALSE)

str(wt)
```



# Test mutational model


## Adding frustration to the LFENM mutational model

Note that v_min changes from wt to mut, but g_entropy doesn't, because the "lfenm" model assumes $K_{mut} = K_{wt}$.
Adding or not frustrations has no effect, because frustration affects the model via $K$, and `wt` is, by construction, fully relaxed.


```{r}
set_wt <- function(pdb, frustrated, node = "sc", model = "ming_wall", d_max = 10.5) {
  set_enm(pdb, node, model, d_max, frustrated)
}

get_mut <- function(wt, site_mut, mutation) {
  get_mutant_site(wt, site_mut, mutation, mut_model = "lfenm", mut_dl_sigma = 0.3, mut_sd_min = 2)
}

tibble(prot = c("mut_nofrust", "mut_frust"),
       site_mut = rep(80, 2),
       mutation = c(1, 1),
       pdb = lst(pdb),
       frustrated = c(F, T)) %>%
  mutate(wt = map2(pdb, frustrated, set_wt)) %>%
  mutate(mut = pmap(list(wt, site_mut, mutation), get_mut)) %>%
  mutate(v_min_wt = map_dbl(wt, enm_v_min),
         v_min_mut = map_dbl(mut, enm_v_min),
         g_entropy_wt = map_dbl(wt, enm_g_entropy, beta = beta_boltzmann()),
         g_entropy_mut = map_dbl(wt, enm_g_entropy, beta = beta_boltzmann())) %>%
  select(frustrated, starts_with("v_min"), starts_with("g_"))
```



## Structural response along sites
```{r}
site = 80
wt <- set_enm(pdb, node = "sc", model = "ming_wall", d_max = 10.5, frustrated = F)
mut <- get_mutant_site(wt, 80, mutation = 1, "lfenm", 0.3, 10.5)
response_site_plot(wt, mut)

```

## Structural response along normal modes

```{r}
response_nm_plot(wt, mut)
```
